Learning a functional distance between organizations with unsupervised embeddings of scientific mobility

### 2020-08-24 paper/current/main.tex

Here, we propose using a neural embedding model that draws from real-world mobility trajectories can encode the complex landscape of human mobility into a dense, continuous, and meaningful vector-space representation, from which not only a meaningful \textit{functional distance} between places but also relationship based on culture, language, and even prestige along with geographic relationship can be probed. We leverage a massive dataset that documents scientific mobility captured in bibliometric records. Scientific mobility, which is central driver of the increasingly-globalized scientific enterprise~\autocite{czaika2018globalisation, altbach2004globalization, box2008competition}, being strongly related to innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation, armano2017innovation}, impact~\autocite{sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}, collaboration~\autocite{rodrigues2016mobility}, and the diffusion of knowledge~\autocite{braunerhjelm2020labor, azoulay2011diffusion, morgan2018prestige}, is not only an important topic on science of science  but also ideal for our study thanks to its richness and well-known structural properties such as the centrality to scientific-advance country, and the strong prestige hierarchy~\autocite{auriol2010careers, clauset2015hierarchy, deville2014career}. In spite of the importance of the problem, understandings of scientific mobility have been limited by the sheer scope and complexity of the phenomenon~\autocite{robinson2019mobility, vannoorden2012mobility, deville2014career, ackers2008mobility, scott2015dynamics}, being further confounded by the diminishing role of geography in shaping the landscape of scientific mobility.

Here, each trajectory is operationalized using a researcher's organizational affiliations derived from their publication histories indexed in a comprehensive bibliographic database (see Methods).
The most fine-grained geographical unit in this data is the organization, such as a ``university'', ``research institute'', ``business'', or ``government agency''.
As a scientist's career progresses, they move between organizations or pick up additional (simultaneous) affiliations forming \textit{affiliation trajectories} (Fig.~\ref{fig:methods:mobility_sentence}).
Thus, the trajectories encode both migration and co-affiliation---the holding of multiple simultaneous co-affiliations involving the sharing of time and capital between locations---that is typical of scientific mobility~\autocite{rodrigues2016mobility, markova2016synchronous, sugimoto2017mostimpact} (see \SI).
We use the trajectories of more than three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database.

A vector-space embedding of organizations is learned by using affiliation trajectories as input to the standard skip-gram with negative sampling, or \textit{word2vec} neural-network architecture (see Methods).
This neural embedding model, originally designed for learning language models~\autocite{mikolov2013word2vec}, has been making breakthroughs by revealing novel insights into texts~\autocite{linzhuo2020hyperbolic, tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2017gendered} and networks~\autocite{perozzi2014deepwalk, grover2016node2vec}.
They are also computationally efficient, robust to noise, and can encode relations between entities as geomemetric relationships in the vector space~\autocite{levy2014neural, nakandala2017gendered, kozlowski2018geometry, an2018semaxis}.
In the organization embedding, each organization is encoded into a single vector, and vectors relate to one another based on the likelihood of organizations appearing adjacent to one another in the same affiliation trajectory.
The functional distance between two organizations is calculated as the cosine distance between their vectors, termed the \textit{embedding distance}.

@@

To assess the performance of the embedding distance relative to geographic distance, we evaluate their correlation with real-world mobility data and their performance in predicting real-world mobility flows as parameters to the gravity model~\autocite{zipf1946gravity, barbosa2018human}.
The Gravity Model is a widely used mobility model~\autocite{curiel2018citygravity, lewer2008immigrationgravity, jung2008highwaygravity, hong2016busgravity, truscott2012epidemicgravity, xia2005measlesgravity} that calculates the \textit{expected flux}, $\hat{T}_{ij}$, of peoples between locations(organizations) based on their populations and distance:

\begin{equation}
	\label{eq:gravity_basic}
	\hat{T_{ij}} = Cm_{i}m_{j}f(r_{ij}),
\end{equation}
where $m_{i}$ is the population of mobile and non-mobile authors who affiliated with institution $i$, $f(r_{ij})$ is a decay function with respect to distance between organizations, and $C$ is a constant estimated from data (see Data and Methods). 
$\hat{T}_{ij}$, which is often referred to as ``expected flux''~\autocite{simini2012universal}, is the expected frequency of the co-occurrence of organizations $i$ and $j$ in the trajectory of affiliations in the gravity model.
The gravity model dictates that the expected flow, $\hat{T}_{ij}$, is symmetric (\ie $\hat{T}_{ij }= \hat{T}_{ji})$, is proportional to the organizations' sizes, $\hat{T}_{ij} \propto m_{i} m_{j}$, and decays as a function of their distance, $f(r_{ij})$. 
We define the distance function in terms of either the geographic distance between organizations or their embedding distance.
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).

@@

\subsection*{Data}
%% Main dataset
% Do we need to talk about how we define mobility in relation to past papers?
We source co-affiliation trajectories of authors from the Web of Science database hosted by the Center for Science and Technology Studies at Leiden University.
Trajectories are constructed from author affiliations listed on the byline of publications for a given author.
Given the limitations of author-name disambiguation, we limit our analyses to papers published after 2008, when the Web of Science began providing full names and institutional affiliations~\autocite{caron2014disambiguation} that improved disambiguation (see \SI).
This yields 33,934,672 author-affiliation combinations representing 12,963,792 authors.
Each author-affiliation combination is associated with the publication year and an ID linking it to one of 8,661 disambiguated organizational affiliations (see \SI for more detail).
Trajectories are represented as the list of author-affiliation combinations, ordered by year of publication, and randomly ordered for combinations within the same year.

>>>>

Here, we propose using a neural embedding model that draws from real-world mobility trajectories can encode the complex landscape of human mobility into a dense, continuous, and meaningful vector-space representation, from which not only a meaningful \textit{functional distance} between places but also relationship based on culture, language, and even prestige along with geographic relationship can be probed. We leverage a massive dataset that documents scientific mobility captured in bibliometric records. Scientific mobility, which is central driver of the globalized scientific enterprise~\autocite{czaika2018globalisation, altbach2004globalization, box2008competition} and strongly related to innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation, armano2017innovation}, impact~\autocite{sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}, collaboration~\autocite{rodrigues2016mobility}, and the diffusion of knowledge~\autocite{braunerhjelm2020labor, azoulay2011diffusion, morgan2018prestige}, is not only an important topic on science of science  but also ideal for our study thanks to its richness and well-known structural properties such as the centrality to scientific-advance country, and the strong prestige hierarchy~\autocite{auriol2010careers, clauset2015hierarchy, deville2014career}. In spite of the its importance, understandings of scientific mobility have been limited by the sheer scope and complexity of the phenomenon~\autocite{robinson2019mobility, vannoorden2012mobility, deville2014career, ackers2008mobility, scott2015dynamics}, being further confounded by the diminishing role of geography in shaping the landscape of scientific mobility.

For the purpose, we use the trajectories of more than three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database.
Each trajectory is operationalized using a researcher's organizational affiliations derived from their publication histories indexed in a comprehensive bibliographic database (see Methods).
As a scientist's career progresses, they move between organizations or pick up additional (simultaneous) affiliations forming \textit{affiliation trajectories} (Fig.~\ref{fig:methods:mobility_sentence}).
Thus, the trajectories encode both migration and co-affiliation---the holding of multiple simultaneous co-affiliations involving the sharing of time and capital between locations---that is typical of scientific mobility~\autocite{rodrigues2016mobility, markova2016synchronous, sugimoto2017mostimpact} (see \SI).

A vector-space embedding of organizations is learned by using affiliation trajectories as input to the standard skip-gram negative sampling, or \textit{word2vec} neural-network architecture (see Methods).
This neural embedding model, originally designed for learning language models~\autocite{mikolov2013word2vec}, has been making breakthroughs by revealing novel insights into texts~\autocite{linzhuo2020hyperbolic, tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2017gendered} and networks~\autocite{perozzi2014deepwalk, grover2016node2vec}.
They are also computationally efficient, robust to noise, and can encode relations between entities as geometric relationships in the vector space~\autocite{levy2014neural, nakandala2017gendered, kozlowski2018geometry, an2018semaxis}.
As a result, each organization is encoded into a single vector representation, and vectors relate to one another based on the likelihood of organizations appearing adjacent to one another in the same affiliation trajectory.

@@

To validate our approach, we evaluate the quality of result vector representations with their performance in predicting real-world mobility flows as parameters to the gravity model~\autocite{zipf1946gravity, barbosa2018human}.
The Gravity Model is a widely used mobility model~\autocite{curiel2018citygravity, lewer2008immigrationgravity, jung2008highwaygravity, hong2016busgravity, truscott2012epidemicgravity, xia2005measlesgravity} that calculates the \textit{expected flux}, $\hat{T}_{ij}$, of peoples between locations (organizations) based on their populations and distance:

\begin{equation}
	\label{eq:gravity_basic}
	\hat{T_{ij}} = Cm_{i}m_{j}f(r_{ij}),
\end{equation}
where $m_{i}$ is the population of mobile and non-mobile authors who affiliated with institution $i$, $f(r_{ij})$ is a decay function with respect to distance between organizations, and $C$ is a constant estimated from data (see Data and Methods). 
$\hat{T}_{ij}$, which is often referred to as ``expected flux''~\autocite{simini2012universal}, is the expected frequency of the co-occurrence of organizations $i$ and $j$ in the trajectory of affiliations in the gravity model.
The gravity model dictates that the expected flow, $\hat{T}_{ij}$, is symmetric (\ie $\hat{T}_{ij }= \hat{T}_{ji})$, is proportional to the organizations' sizes, $\hat{T}_{ij} \propto m_{i} m_{j}$, and decays as a function of their distance, $f(r_{ij})$. 
We define the distance function in terms of either the geographic distance between organizations or their functional distance in the vector space, which is calculated as the cosine distance between their vectors, termed the \textit{embedding distance}. For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).

@@

\subsection*{Data}
%% Main dataset
% Do we need to talk about how we define mobility in relation to past papers?
We source co-affiliation trajectories of authors from the Web of Science database hosted by the Center for Science and Technology Studies at Leiden University.
Trajectories are constructed from author affiliations listed on the byline of publications for a given author.
Given the limitations of author-name disambiguation, we limit our analyses to papers published after 2008, when the Web of Science began providing full names and institutional affiliations~\autocite{caron2014disambiguation} that improved disambiguation (see \SI).
This yields 33,934,672 author-affiliation combinations representing 12,963,792 authors.
Each author-affiliation combination is associated with the publication year and an ID linking it to one of 8,661 disambiguated organizational affiliations (see \SI for more detail).
Trajectories are represented as the list of author-affiliation combinations, ordered by year of publication, and randomly ordered for combinations within the same year. The most fine-grained geographical unit in this data is the organization, such as a ``university'', ``research institute'', ``business'', or ``government agency''.
